---
title: "DA5020.A6.Nithya.Sarabudla"
author: "nithyasarabudla"
date: "2023-10-21"
output: pdf_document
---

## Importing libraries
```{r}
library(RSQLite)
library(ggplot2)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Bonus part 2 
What is the average runtime for the thriller movie genre.

```{r}
db_path <- "/Users/nithyasarabudla/Downloads/SQlite/imdb.db"  # Replace with the actual file path
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
query <- "SELECT AVG(Runtime) 
              FROM movie_info
              WHERE Genre LIKE '%Thriller%'"

result <- dbGetQuery(con, query)
result

```

## Question 1 

### Part 1 

Create a table named director_info using SQLite; the columns are: Director_ID, and
Director_Name. The Director_ID should be the primary key.
```sql
CREATE TABLE director_info (
    Director_ID INTEGER PRIMARY KEY,
    Director_Name TEXT
)
```
### Part 2

Import the entire data from the CSV file into the director_info table using the SQLite .import
command (see helpful resources below). Verify that your data was imported correctly

```sql
.import "/Users/nithyasarabudla/Downloads/SQlite/directors.csv" director_info"
```
## Question 2
### Conneting to Database

```{r}
db_path <- "/Users/nithyasarabudla/Downloads/SQlite/imdb.db"  # Replace with the actual file path
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
```


### Part 1
Count the number of rows in the movie_info and director_info tables.
```{r}
# Execute a SELECT query
result1 <- dbGetQuery(con, "SELECT count(*) as movie_count FROM movie_info")
result2 <- dbGetQuery(con, "SELECT count(*) as director_info FROM director_info")

# Print the result
print(result1)
print(result2)
```


### Part 2
How many movies were released between 2010 and 2020 (inclusive)? Visualize the results.
```{r}
# Execute a SELECT query
query <- "SELECT Release_Year, COUNT(*) as num_movies FROM movie_info where Release_Year between 2010 AND 2020 GROUP BY Release_Year"
result <- dbGetQuery(con, query)
result

ggplot(result, aes(x = Release_Year, y = num_movies)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of Movies Released Each Year", x = "Release Year", y = "Number of Movies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Part 3
What is the minimum, average and maximum ratings for “Action” movies. Ensure that you
query the genre using wild cards.
```{r}
min_rating <- dbGetQuery(con, "SELECT MIN(IMDB_Rating) as min_rating
FROM movie_info
WHERE upper(Genre) LIKE '%ACTION%'")
  
max_rating <- dbGetQuery(con, "Select max(IMDB_Rating) from movie_info where upper(Genre) LIKE '%ACTION%' ")
  
avg_rating <- dbGetQuery(con, "Select AVG(IMDB_Rating) from movie_info where upper(Genre) LIKE '%ACTION%' ")


print(min_rating)
print(max_rating)
print(avg_rating)
  
```
### Part 4
What are the 25 highest-grossing movies within the dataset? Display the title, genre and gros
```{r}
query <- "SELECT Series_Title, Genre, Gross
FROM movie_info
WHERE Gross IS NOT 'NA'
ORDER BY Gross DESC
LIMIT 25"

result <- dbGetQuery(con,query)

print(result)
  
```
### Part 5
Which directors have the highest-grossing movies. Display the director name and the total
gross. Ensure that you join the necessary tables. Visualize the results using a Bar chart

```{r}
query <- "SELECT d.Director_Name, SUM(m.Gross) as Total_Gross
FROM movie_info m
JOIN director_info d ON m.Director_ID = d.Director_ID
GROUP BY d.Director_Name ORDER BY Total_Gross DESC
LIMIT 25 ";

result <- dbGetQuery(con,query)

ggplot(result, aes(x = Director_Name, y = Total_Gross)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Gross Earning Plot", x = "Director Name", y = "Gross Earnings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(result)
  
```
### Part 6 
Create a function called verifyDirector() that takes a director name as its argument, and
queries the database to check if the director exists. Your function should display a message to notify
the user if the director was found or not.

```{r}
verifyDir <- function(director_name) {

  # Prepare the SQL query to check if the director exists
  query <- sprintf("SELECT COUNT(*) as count FROM director_info WHERE Director_Name = '%s'", director_name)
  
  # Execute the query
  result <- dbGetQuery(con, query)
  
  
  # Check if the director exists (count > 0)
  if (result$count > 0) {
    print("DIRECTOR FOUND IN DATABSE !")
  } else {
    print("DIRECTOR NOT FOUND IN DATABSE !")
  }
}

# Director present - check
verifyDir("Anthony Russo")

#Director not present - check
verifyDir("Nithya S")

```


