---
title: "DA5020"
author: "nithyasarabudla"
date: "2023-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1 
```{r}
library(XML)
library(RCurl)
library(tidyverse)
library(dplyr)
library(lubridate)
library(httr)

# Load the XML data directly from the URL below into a data frame 
url <- "http://www.senate.gov/general/contact_information/senators_cfm.xml"
xml <- xmlParse(rawToChar(GET(url)$content))
xml_df <- xmlToDataFrame(xml)

# Displaying the dimensions of the data 
dim(xml_df)
```

# Question 2 
```{r}
# Constructing regular expression (regex) to extract only the first, last names, and party (D,R,I) of each senator
senator <-str_extract(xml_df$first_name, "^[A-Z][a-z]+")
# Mutate the 'first_name' column in the 'xml_df' data frame with the extracted first names
senate <-xml_df %>%
  mutate(first_name = senator)
# Create a new data frame 'updated_df' with the updated 'first_name' and 'last_name' columns,
# as well as the 'party' column from the 'senate' data frame
updated_df<- data_frame(First_name = senate$first_name, Last_name = senate$last_name, Party = senate$party)
# Convert 'updated_df' to a standard data frame and display the result
as.data.frame(updated_df)
```

# Question 3 
```{r}
# Create a function called senatorsByState() 
senatorsByState <- function(x){
# Filter senators by the state abbreviation and select relevant columns
    state_senate <- senate %>%
    filter(state == x) %>%
    select(state,first_name, last_name, party) %>%
    group_by(state)
# Create a message indicating the state abbreviation
  state_message <- paste("The senators for",(x),"are: ")
 # Create a string containing senator information (first name, last name, and party)
  senator_info <- paste(state_senate$first_name, state_senate$last_name, ',', recode(state_senate$party, "D" = "Democratic Party", "R" = "Republican Party", "I" = "Independent Party"))
# Combine the senator information using 'and' as a separator
  senators_combined <- str_c(senator_info, collapse = " and ")
 # Combine the state message and the senator information
  result_message <- str_c(state_message, senators_combined)
  return(result_message)
}
# Example
senatorsByState("MA")

```

# Question 4 
```{r}
# Load the data from the CSV file 
unemployment_data <- read.delim2("Ratio Of Female To Male Youth Unemployment Rate .csv", header=TRUE, sep = ",", skip=4)
# Create a tibble with country names and codes
country_info <- tibble(
  Country_Name = unemployment_data$Country.Name,
  Country_Code = unemployment_data$Country.Code
)
# Remove duplicate country entries
country_name <- country_info %>%
  distinct()
country_name

year_Value <- unemployment_data%>% 
# Pivot the data to long format for years X1960 to X2020
  pivot_longer(c('X1960':'X2020'), names_to = "Year", values_to = "Value")
# Create a new data frame for indicator data
indicator_Data = data.frame(unemployment_data$Country.Code, year_Value$Year, year_Value$Value)
colnames(indicator_Data)[1] = "Country_Code"
colnames(indicator_Data)[2] = "Year"
colnames(indicator_Data)[3] = "Value"
# Convert the indicator data to a tibble
indicator_data <- as_tibble(indicator_Data)
indicator_data
```

# Question 5 
```{r}
Continent_Name <- c("North America", "North America", "North America", "North America", "North America", "Asia", "Asia", "Asia", "Asia", "Asia", "Middle East", "Middle East", "Middle East", "Middle East", "Middle East")
Country_Name <- c("United States", "Canada", "Mexico", "Greenland", "Cuba", "India", "China", 
"Pakistan", "Afghanistan", "Bangladesh", "Saudi Arabia", "Yemen, Rep.", "Israel", "Iran, Islamic Rep.", "Jordan")
continent_df <- data.frame(Continent_Name, Country_Name)

join_df <- continent_df %>%
  left_join(country_info, by = "Country_Name") %>%
  left_join(indicator_data, by = "Country_Code")


join_df %>%
  filter(Year %in% c("X2001", "X2002", "X2003", "X2004", "X2005", "X2006", "X2007", "X2008", "X2009", "X2010", "X2011", "X2012", "X2013", "X2014", "X2015", "X2016", "X2017", "X2018", "X2019", "X2020")) %>%
  ggplot(aes(Year, Value)) +
  geom_line(aes(group = Country_Name, color = Country_Name)) +
  facet_wrap(~Continent_Name, nrow = 3)

```

In Asia most of the countries in this segment show fluctuating values over the years. The china and India seem to be the most dominant in terms of magnitude.

In Middle Eastern countries the ratio Of Female To Male Youth Unemployment Rate was the highest in Yemen in 2002, Saudi Arabia in 2003, Iran in 2006, Israel in 2008.

In North America the trends for the countries in this region are fluctuating, with the green line  representing the United States being notably prominent.
