---
title: "Assignment 1"
author: "nithyasarabudla"
date: "2023-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1 
## Data frame contains different data types and data displayed in the table format.data.frame() function is used to create a data frame
### Example 
```{r }
Market <- data.frame ( 
  fruits = c("apple", "banana", "grapes"),
  cost = c(10,12,14),
  weight_kgs = c(2,4,5) )
# printing the data frame 
Market
```

# Question 2

```{r}
df <- data.frame(cars = c("Truck", "Car", "SUV"),
                  mpg = c(11, 30, 24),
                  cost = c(45000,25000,35000) )
df
```
# Question 2a
## Select row 1 in column 3
```{r}
df[1,3]
```
## Select rows 1 through 3
```{r}
df[1:3,]
```
## Select the last column
```{r}
df[,3]
```

# Question 3
## using the build in data set mtcars
```{r}
data(mtcars)
```
## Display the first 3 rows of the dataset and display the last 5 rows.
```{r}
head(mtcars,3)
```

```{r}
tail(mtcars,5)
```

# Question 4

```{r}
var <- names(sapply(mtcars,typeof))
# categorical variables
var[c(2,8)]
# continuous variables
var[c(1,3)]
```

# Question 5 

```{r }
library(ggplot2)
ggplot(mtcars,aes(x=disp, y=mpg)) + geom_point()
```
## The empty chat is displayed with the x and y axis of mpg and disp but the points are not marked. The function geom_point() is missing.It adds a layer of points to your plot, which creates a scatterplot.

# Question 6
```{r}
data(mpg)
ggplot(mpg,aes(x=displ, y=hwy , color=cyl)) + geom_point()
```
## The chart indicates that as the number of cylinders increases, there is a significant decrease in highway miles per gallon. Additionally, the number of cylinders increases, the engine displacement also increases.

# Question 7
```{r}
ggplot(mpg,aes(x=displ, y=hwy , color=cyl)) + geom_point() + facet_wrap(~drv)
```
## In comparison to both 4wd and front-wheel drive, rear-wheel drive tends to have a larger engine displacement, ranging from 3 to 7 liters, even though it utilizes fewer cylinders. whereas 4wd and front-wheel drive has almost similar highway miles per gallon but lesser engine displacement.
 
 
