---
title: "DA5020.A2.Nithya.Sarabudla"
author: "nithyasarabudla"
date: "2023-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
data("msleep")
```
## question 1 
(2)  "msleep" dataset contains 83 rows and 11 variables. The dataset contains the common name, taxonomic rank, vore, order,the conservation status of the mammal,  sleep times and weights for a set of mammals. There is some missing data in the msleep dataset.

## question 2 
```{r}
# Calculating the proportion of each category of vore as a percentage
category_vore <- msleep %>%
  group_by(vore) %>%
  summarize(count = n()) %>%
  mutate(proportion = count / sum(count) * 100)
# visualizing the results using a barchart
ggplot(category_vore, aes(x = vore, y = proportion)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of each category of vore ",
    x = "Vore",
    y = "proportions"
  )
```

In the bargraph herbivores has the highest percentage and insectivores has the lowest percentage. 

## question 3
```{r}
# Filtering the data to extract data for just omnivores
filtered_data = filter(msleep, vore=="omni")
# calculating the mean sleep_total for that group
 summarize(filtered_data, mean( sleep_total))
```

## question 4

```{r}
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "bodywt",
    y = "brainwt",
    title = "  Scatterplot of bodywt vs brainwt"
  )
```

The graph shows the positive correlation.

## question 5
```{r}
# Pearson correlation coefficient
pearson_cor<- cor(msleep$bodywt, msleep$brainwt, method = "pearson", use="complete.obs")
pearson_cor
```

From the results bodywt and brainwt are highly correlated. The results support the original assumptions from question 4

## question 6 
```{r}
mean_sleep <- mean(msleep$sleep_total)
median_sleep <- median(msleep$sleep_total)
sd <- sd(msleep$sleep_total)
msleep$z <- abs((mean_sleep-msleep$sleep_total)/sd)
outliners <- msleep[which(msleep$z >1.5),]
outliners_st <- outliners[, c("name", "sleep_total")]
print(outliners_st)
```




