---
title: "DA5020.A3.Nithya.Sarabudla"
author: "nithyasarabudla"
date: "2023-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

# question 2
```{r}
#Load the data from the CSV file into the tripdata_df dataframe
tripdata_df <-read.csv("2018_Green_Taxi_Trip_Data.csv")

# Display the dimensions (number of rows and columns) of the dataframe
dim(tripdata_df)
# Provide a summary of the dataframe's structure and data types
glimpse(tripdata_df) 
# Generate a summary of the statistics for the dataframe's numeric columns
summary(tripdata_df)

# Count the occurrences of missing values in the dataframe
count(tripdata_df, "NA")
```

# question 3
```{r}
# Calculating the frequency of missing values in each column
missing_values <- colSums(is.na(tripdata_df))
print(missing_values)
```
This code helps us see how many times data is missing in our dataset. It helps us find columns with a lot of missing data.

```{r}
head(tripdata_df$lpep_dropoff_datetime)
glimpse(tripdata_df$lpep_pickup_datetime)
```
From the above it's noticeable that in the 'lpep_dropoff_datetime' and 'lpep_pickup_datetime' variables, the date corresponds to January. However, we downloaded the CSV file intending to cover the month of February, and the time format is also inconsistent.

```{r}
summary(tripdata_df$tip_amount)
summary(tripdata_df$trip_duration_seconds)
```
In the above code summary statistics and visualizations is used,to identify outliers or invalid values in the tip_amount and  trip_duration_seconds

# question 4 
```{r}
data <- as.data.frame(tripdata_df$trip_distance) 
histogram <- ggplot(data, aes(x = tripdata_df$trip_distance)) +
  geom_histogram()+
   scale_y_log10()+
  labs(x = "Trip distance in miles ", title= "histogram showing trip_distance")
histogram
```

The data is skewed,the histogram indicates that the vast majority of trips are less than 75 miles in distance, with a significant portion even falling between 0 and 25 miles.

# question 5

```{r}

# mean and standard deviation of trip_distance
mean_trip_distance <- mean(tripdata_df$trip_distance)
sd_trip_distance <- sd(tripdata_df$trip_distance)

# mean and standard deviation of tip_amount 
mean_tip_amount <- mean(tripdata_df$tip_amount)
sd_tip_amount <- sd(tripdata_df$tip_amount)

# outlier threshold oftip_amount and trip_distance
threshold_tip_amount <- 3 * sd_tip_amount
threshold_trip_distance <- 3 * sd_trip_distance

# Identifing outliers in trip_distance
outliers_trip_distance <- tripdata_df$trip_distance < (mean_trip_distance - threshold_trip_distance) | tripdata_df$trip_distance > (mean_trip_distance + threshold_trip_distance)

# Identifing outliers in tip_amount 
outliers_tip_amount <- tripdata_df$tip_amount < (mean_tip_amount - threshold_tip_amount) | tripdata_df$tip_amount > (mean_tip_amount + threshold_tip_amount)

# Counting the number of outliers
num_of_tip_amount_outliers <- sum(outliers_tip_amount)
print(num_of_tip_amount_outliers)
num_of_trip_distance_outliers <- sum(outliers_trip_distance)
print(num_of_trip_distance_outliers)

# removing the outlier tip_amount from the data
tripdata_df <- tripdata_df[!outliers_tip_amount, ]

```
From the above code outliers detected in tip_amount and trip_distance variable are 14444 and 20953.

# question 6
```{r}
# Filter trip_distance and removing the outliers
filtered_data <- tripdata_df[!outliers_trip_distance, ]

# Create a bar chart of trip_distance by payment_type
ggplot(filtered_data, aes(x = payment_type, y=frequency(payment_type))) +geom_col(color="blue") +ylab("counts") +xlab("payment type") + ggtitle("The frequency of the payment types") 
```


# question 7

We can handle the missing data by 

1.Filling in the missing values with estimated or calculated values with the mean/median/most frequent value of the respective column.

2.Deleting the rows that contains missing values.

There are only three missing values in the "trip_type" column out of a total of 1,048,575 values. While one solution is to remove these three rows, it may not significantly affect the dataset since it's a very small fraction. However, considering that nearly 99% of the values are equal to 1, we could alternatively choose to impute these three missing values with the value 1.
```{r}
# checking the count of the missing values in "trip_type" column
sum(is.na(tripdata_df$trip_type))
# "trip_type" column by replacing the NA values with the value 1
tripdata_df$trip_type[is.na(tripdata_df$trip_type)] <- 1
```

In the Column "ehail_fee" the  entire column is of NA values. Therefore the entire column can be dropped.
```{r}
# removing the entire column of NA values in ehail_fee
tripdata_df <- within(tripdata_df, rm('ehail_fee'))
```

