---
title: "DA5020.A7.Nithya.Sarabudla"
author: "nithyasarabudla"
date: "2023-10-25"
output:
  pdf_document: default
  html_document: default
  
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## loading the packages 
```{r}
library(rvest)
library(dplyr)
library(ggplot2)
library (tidyverse)
```

## extract the tabular data on the “Percent of 
## population living on less than $1.15, $2.60 and $3.85 a day” from the Wikipedia page.
# Question 1 
# PART 1
##  Scrape the data from the webpage and extract the following fields: Country, < $1.15, <
## $2.60, < $3.85, Year and Continent. Prepare the data for analysis and ensure that the columns have
## meaningful names.

```{r}
# Read the HTML page from the given URL
page = read_html("https://en.wikipedia.org/w/index.php?title=List_of_sovereign_states_by_percentage_of_population_living_in_poverty&diff=prev&oldid=1177591928")

# Extract the third table from the HTML page
table <- html_table(page)[3]
table
# Convert the table to a data frame
df <- data.frame(table)

# Rename the columns
colnames(df) <- c("Country", "less_than_1.15", "less_than_2.60", "less_than_3.85", "Year", "Continent")

# Clean and convert the "less_than_1.15" column by removing any trailing characters and converting to numeric
df$less_than_1.15 <- as.array(df$less_than_1.15)
df$less_than_1.15 <- substring(df$less_than_1.15,1,nchar(df$less_than_1.15)-1)
df$less_than_1.15 <- as.numeric(df$less_than_1.15)

# Clean and convert the "less_than_2.60" column by removing any trailing characters and converting to numeric
df$less_than_2.60<- as.array(df$less_than_2.60) 
df$less_than_2.60 <- substring(df$less_than_2.60,1,nchar(df$less_than_2.60)-1) 
df$less_than_2.60 <- as.numeric(df$less_than_2.60)

# Clean and convert the "less_than_3.85" column by removing any trailing characters and converting to numeric
df$less_than_3.85 <- as.array(df$less_than_3.85) 
df$less_than_3.85 <- substring(df$less_than_3.85,1,nchar(df$less_than_3.85)-1) 
df$less_than_3.85<- as.numeric(df$less_than_3.85)

# Display the first few rows of the data frame
head(df)
```

# PART 2
##  Calculate the mean and the standard deviation of the percent of the population living under
## $3.85 per day for each continent. Perform a comparative analysis (i.e. explanation) of the data from
## each continent

```{r}
# Calculate the mean and standard deviation of "less_than_3.85" percentages for each continent
less_than_3.85_df <- df %>% 
  select(Continent, less_than_3.85) %>% 
  group_by(Continent) %>% 
  summarise(Mean = mean(less_than_3.85, na.rm = TRUE), std_dev = sd(less_than_3.85, na.rm = TRUE))

less_than_3.85_df

# Find and display the continent with the lowest mean poverty level
min_mean_continent <- less_than_3.85_df %>%
  filter(Mean == min(Mean)) %>%
  select(Continent) %>%
  pull()
cat("Continent with the lowest mean poverty level:", min_mean_continent, "\n")

# Find and display the continent with the lowest standard deviation
min_std_dev_continent <- less_than_3.85_df %>%
  filter(std_dev == min(std_dev)) %>%
  select(Continent) %>%
  pull()
cat("Continent with the lowest standard deviation:", min_std_dev_continent, "\n")

# Find and display the continent with the highest mean poverty level
max_mean_continent <- less_than_3.85_df %>%
  filter(Mean == max(Mean)) %>%
  select(Continent) %>%
  pull()
cat("Continent with the highest mean poverty level:", max_mean_continent, "\n")

# Find and display the continent with the highest standard deviation
max_std_dev_continent <- less_than_3.85_df %>%
  filter(std_dev == max(std_dev)) %>%
  select(Continent) %>%
  pull()
cat("Continent with the highest standard deviation:", max_std_dev_continent, "\n")

```

# PART 3 
##  What are the 10 countries with the highest percentage of the population having an income
## of less than $3.85 per day? Using a suitable chart, display the country name, the percentage and
## color- code by the Continent. Summarize your findings
```{r}
# Extract the top 10 countries with the highest percentage of income less than $3.85
top_10_countries <- df %>%
  arrange(desc(less_than_3.85)) %>%
  head(10)

# Create a bar plot to visualize the top 10 countries' income percentages
df %>%
  arrange(desc(less_than_3.85)) %>%
  select(Country, less_than_3.85, Continent) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Country, desc(less_than_3.85)), y = less_than_3.85, fill = Continent)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90))

# Summarize your findings
summary(top_10_countries)

```

From the above plot, it's evident that the majority of the countries with the highest percentage of the population living 
on less than $3.85 per day are from Africa. Specifically, out of the top 10 countries in this category, 9 are from Africa,  while only 1 country is from Asia. South Sudan ranks as the top country with the highest percentage of its population
earning less than $3.85 per day. The 10th country on the list with the highest population living on less than $3.85 per day is Mali, also from Africa.

# PART 4 
## Explore the countries with the lowest percentage of the population having an income of
## less than $3.85 per day. What are the 5 countries with the lowest percentage, and how does the
## results compare to the other income groups (i.e. $1.15 and $2.60)?
```{r}
# Arrange the data frame in ascending order of the "less_than_3.85" column
# and select the top 5 rows with the lowest income percentages
df %>% arrange(less_than_3.85) %>% head(5)

```
Among the 5 countries with the lowest population percentage living on less than $3.85 per day, 2 are in Asia (Azerbaijan
the United Arab Emirates) with a 0% population in this category, while the remaining 3 are from Europe (Iceland, Belarus, 
and Finland) with percentages ranging from 0.04% to 0.1%. For the less than $2.60 per day threshold, only Finland has 0.1%,  with the rest at 0%. In all these countries, when considering less than $1.15 per day, the population percentage is 0%.

# PART 5 
## Extract the data for any two continents of your choice. For each continent, visualize the
## percent of the population living on less than $1.90, $3.20 and $5.50 using box plots. Compare and
## contrast the results, while ensuring that you discuss the distribution, skew and any outliers that are
## evident.
```{r}
# Create a box plot to visualize the income distribution in Africa
df %>%
  filter(Continent == "Africa") %>%
  pivot_longer(cols = starts_with("less"), names_to = "Income Level", values_to = "Percentage") %>%
  ggplot(aes(x = `Income Level`, y = Percentage)) +
  geom_boxplot() +
  ggtitle("Box plot of population in Africa") +
  xlab("Income Level") + 
  ylab("% of Population")

# Create a box plot to visualize the income distribution in Europe
df %>%
  filter(Continent == "Europe") %>%
  pivot_longer(cols = starts_with("less"), names_to = "Income Level", values_to = "Percentage") %>%
  ggplot(aes(x = `Income Level`, y = Percentage)) +
  geom_boxplot() +
  ggtitle("Box plot of population in Europe") +
  xlab("Income Level") + 
  ylab("% of Population")
```

## Box Plot for Asia:
Distribution: The box plot for "Asia" shows that the distribution of the population living on less than $1.15 is right skewed, while the distribution for less than $2.60 is slightly left-skewed, and the distribution for less than $3.85 is  strongly left-skewed.

Skew: Right-skewed for less than $1.15, with most people having lower income.
Slightly left-skewed for less than $2.60, where the majority have higher income.
Strongly left-skewed for less than $3.85, indicating that the majority have even higher income levels.

Outliers: There are 4 outliers in less_than_3.85 means there are a few places or individuals in Asia with much lower income compared to most people. 

## Box Plot for Europe :
In Europe, the income distribution for less than $1.15, less than $2.60, and less than $3.85 is strongly right-skewed.
This indicates that the majority of the population in Europe has higher income levels, with a long tail of individuals who have lower income. There are a few outliers in the three income levels, which represent exceptional cases of individuals or regions with significantly lower income than the majority of the population. 
